\documentclass[12pt]{report}
\usepackage{caption}
\captionsetup{font=normalsize}

\usepackage{algorithm}


% Add additional packages here
\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{ amssymb }
\usepackage{array}
\usepackage{wrapfig}
\usepackage{wrapfig}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage{float}
\usepackage{tabu}
\usepackage{longtable}
\usepackage[nottoc]{tocbibind}
\usepackage{tabularx}
% Include any extra LaTeX packages required
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{times}



\nocite{*}
\makeatletter
\DeclareRobustCommand{\textsupsub}[2]{{%
  \m@th\ensuremath{%
    ^{\mbox{\fontsize\sf@size\z@#1}}%
    _{\mbox{\fontsize\sf@size\z@#2}}%
  }%
}}
\makeatother
\newbibmacro*{volume+number}{%
  \printfield{volume}%
  \setunit*{(}{}
  \printfield{number}%
  \setunit*{)}{}
}\usepackage{float}
\begin{document}
\pagenumbering{roman}

\begin{titlepage}
\input{titilepage}

\end{titlepage}
\newpage

%Declaration part(Second Page)
\input{Declaration}

%Certificate part(Third Page)
\begin{titlepage}
\input{certificate}
\end{titlepage}
\newpage

%Appproval

\begin{titlepage}
\input{Approval List}
\end{titlepage}
\newpage


%Acknowledgement part(Fourth Page)



\input{Acknowledgement}
% End of the Acknowledgements
%% ----------------------------------------------------------------
\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Modern smart grids operate as complex cyber-physical systems where electrical infrastructure, communication networks, and control mechanisms are tightly integrated. This integration improves efficiency and observability but also increases exposure to coordinated cyber-physical attacks. Static threshold monitoring and periodic audits cannot address the dynamic, distributed nature of such environments. This project presents a SmartGrid AI Audit Framework that provides continuous monitoring, explainable anomaly detection, and intelligent audit decision support within a single operational platform.

\par \vspace{0.35cm}
The system runs in two workspaces: \textit{Experiment Running} for simulation and benchmarking, and \textit{Rapid SCADA Live} for real-time monitoring over a 100-agent grid of generators, substations, PMUs, and breakers. A PowerShell bridge acquires SCADA telemetry through the Web API and forwards it to a FastAPI backend. The detection pipeline combines statistical deviation scoring with a dual-branch LSTM temporal detector, pre-trained on the UCI Electrical Grid Stability Dataset (physical branch) and the UNSW-NB15 Network Intrusion Dataset (cyber branch), using a two-stage fusion: Stage~1 fuses the physical and cyber LSTM branches ($w_{\text{grid}}=0.58$, $w_{\text{network}}=0.42$); Stage~2 fuses the LSTM output with deviation scoring ($w_{\text{dev}}=0.48$, $w_{\text{prob}}=0.52$). A multi-layer detector architecture adds a calibrated threshold layer, a temporal accumulator, and four attack-specific detectors for FDI (CUSUM with shape gate, $k=0.50$, $h=15.0$), DoS (weighted rule), MITM (integrity-jump), and physical faults (signature template), combined by an OR-with-precedence rule and a cross-layer corroboration gate. A hybrid audit scheduler combining Q-learning over a 4D state space (300 states) with gradient-based cost optimisation determines escalation actions, and a per-feature XAI module provides operator-facing explanations.

\par \vspace{0.35cm}
At N=100, in the system level configuration with audit protection active, the framework achieves 93.85\% detection accuracy, 6.18\% false positive rate, 98.24\% recall, 57.28\% cost efficiency, and 100\% audit coverage, averaged across five seeds. Measured at the detector level in evaluation mode, the per attack true positive rates are 99.0\% for FDI, 78.8\% for DoS, 72.6\% for MITM, and 98.4\% for physical faults at N=100, where the fault class was raised from 8.1\% by the new signature detector. Multi-seed evaluation (seeds 42 to 46) shows a standard deviation of 0.33\% in accuracy, confirming result stability. Scalability experiments at N=200 and N=500 show that detection accuracy and false positive rate stay nearly constant while the audit budget becomes the limiting resource at the largest size. The SCADA pipeline is fully operational; current telemetry uses Rapid SCADA calculated channels rather than physical field devices.

\par \vspace{0.35cm}
\textit{Keywords: Smart Grid, Cyber-Physical Systems, Anomaly Detection, Audit Intelligence, Rapid SCADA, Explainable AI, Multi-Layer Detection, LSTM, CUSUM, Q-Learning, Hybrid Scheduling, UCI Dataset, UNSW-NB15, Cybersecurity}

%Contents part(Page 5)
\tableofcontents

\listoffigures
\listoftables
\newpage
%Introduction part(Page 6)
\pagenumbering{arabic}

\chapter{Introduction}
\input{Chapter1}

\chapter{Literature Survey}
 \input{Chapter2}

\chapter{Proposed System Architecture \& Design}
 \input{Chapter3}

\chapter{Implementations }
\input{Chapter4}

\chapter{Base Paper And Its Limitation}
\input{Chapter5}

\chapter{Testing}
\input{Chapter6}

\chapter{Result \&  Analysis}
\input{Chapter7}

\chapter*{Conclusion}
\input{Conclusion}

 \chapter*{Project Scheduling and Planning}
 \input{Chart}

\printbibliography
\end{document}
